---
title: "Using Linear Programming"
author: "Tinniam V Ganesh"
date: "September 28, 2017"
output: html_document
---

# Using Linear Programming (LP) for optimizing the bowling and batting lineup
In my recent post [My travels through the realms of Data Science, Machine Learning, Deep Learning and (AI)](https://gigadom.wordpress.com/2017/09/11/my-travels-through-the-realms-of-data-science-machine-learning-deep-learning-and-ai/),I had recounted my journey in the domains of  of Data Science, Machine Learning (ML),
and more recently Deep Learning (DL) all of which are useful while analyzing data. Of late, I have come 
to the realization that there are many facets to data. And to glean insights from data, Data Science, ML and DL alone are not sufficient and one needs to also have a good handle on linear programming and optimization. My colleague at IBM Research also concurred with this view and told me he had arrived at this conclusion several years ago. While ML & DL are useful and interesting to make inferences and predictions of outputs from input variables, optimization computes the choice of input which result in  maximum or minimum. So I made a small course correction and started on a course from India's own NPTEL [Introduction to Linear Programming](http://freevideolectures.com/Course/2365/Fundamentals-of-Operations-Research) by Prof G. Srinivasan of IIT Madras. The lectures are delivered with remarkable clarity by the Prof and I am just about halfway through the course (each lecture is of 50-55 min duration) when I decided that I needed to try to formulate and solve some real world Linear Programming problem. 

As usual I turned towards cricket for some appropriate situations and sure enough it was there in the open. For this LP formulation I take International T20 and IPL, though International ODI also will work equally well.

Assuming there are 3 bowlers - $$bwlr_{1},bwlr_{2},bwlr_{3}$$
and there are 4 batsmen - $$bman_{1},bman_{2},bman_{3},bman_{4}$$

Let  the economy rate $$er_{ij}$$ be the Economy Rate of the jth bowler to the ith batsman. Also if remaining overs  for the bowlers are  $$o_{1},o_{2},o_{3}$$
and the total number of overs left to be bowled are
$$o_{1}+o_{2}+o_{3} = N$$
then 
How many overs should each bowler bowl, so that the total runs scored by the batsmen are minimum?

## 1. LP Formulation for bowling order
Let  the economy rate $$er_{ij}$$ be the Economy Rate of the jth bowler to the ith batsman.
**Objective function** : Minimize - $$\sum_{i=1}^{i=4}\sum_{j=1}^{i=3}er_{ij}*o_{j}$$
**Constraints** 
Where k is the number overs o, remaining for the jth bowler $$o_{j} <= k_{j}$$ 
and the total number of overs remaining to be bowled is N then -$$\sum o_{j} = N$$ 
Also
$$o_{j} >=0$$ 
The overs that any bowler can bowl can be >=0

## 2. LP Formulation for batting lineup
Let  the strike rate $$sr_{ij}$$ be the Strike Rate of the ith batsman to the jth bowler
**Objective function** : Maximize - $$\sum_{i=1}^{i=4}\sum_{j=1}^{i=3}sr_{ij}*o_{j}$$
**Constraints** 
Where k is the number overs o, remaining for the jth bowler $$o_{j} <= k_{j}$$ 
and the total number of overs remaining to be bowled is N then -$$\sum o_{j} = N$$ 
Also
$$o_{j} >=0$$ 
The overs that any bowler can bowl can be >= 0 or any number that the bowler has already bowled.

For this maximization and minimization problem I used lpSolveAPI.

## 3. LP formulation (Example 1)
Initially I created a test example to ensure that I get the LP formulation and solution correct.
Here the er1=4 and er2=3 and o1 & o2 are the overs bowled by bowlers 1 & 2. Also o1+o2=4
In this example as below

B1	B2	Obj Fun(=4*B1+3*B2)  
1	3	   13  
2	2	   14  
3	1	   15  

```{r message=FALSE }
library(lpSolveAPI)
library(dplyr)
library(knitr)
lprec <- make.lp(0, 2)
a <-lp.control(lprec, sense="min")
set.objfn(lprec, c(4, 3))  # Economy Rate of 4 and 3 for er1 and er2
add.constraint(lprec, c(1, 1), "=",4)  # o1 + o2 =4
add.constraint(lprec, c(1, 0), ">",1)  # o1 > 1
add.constraint(lprec, c(0, 1), ">",1)  # o2 > 1
lprec
b <-solve(lprec)
get.objective(lprec) # 13
get.variables(lprec) # 1    3 
```

**Note 1:** In the above example **13 runs** is the minimum that can be scored and this requires

- o1=1
- o2=3

**Note 2:**The numbers in the columns represent the number of overs that need to be bowled by a bowler to the
corresponding batsman.

## 4. LP formulation (Example 2)
In this formulation there are 2 bowlers and 2 batsmen
o11,o12 are the oves bowled by  bowler 1 to batsmen 1 & 2
and o21, o22 are the overs bowled by bowler 2 to batsmen 1 & 2
er11=4, er12=2,er21=2,er22=5
o11+o12+o21+o22=5

The solution for this manually computed is
B1	B2	B1	B2	Runs  
1	1	1	2	18  
1	2	1	1	15  
2	1	1	1	17  
1	1	2	1	15  

```{r }
lprec <- make.lp(0, 4)
a <-lp.control(lprec, sense="min")
set.objfn(lprec, c(4, 2,2,5))
add.constraint(lprec, c(1, 1,0,0), "<=",8)
add.constraint(lprec, c(0, 0,1,1), "<=",7)
add.constraint(lprec, c(1, 1,1,1), "=",5)
add.constraint(lprec, c(1, 0,0,0), ">",1)
add.constraint(lprec, c(0, 1,0,0), ">",1)
add.constraint(lprec, c(0, 0,1,0), ">",1)
add.constraint(lprec, c(0, 0,0,1), ">",1)
lprec
b<-solve(lprec)
get.objective(lprec) 
get.variables(lprec) 
```

Note: In the above example **15 runs** is the minimum that can be scored and this requires

- o11=1
- o12=2
- o21=1
- o22=1

It is possible to keep the minimum to other values and solves also.

## 5. LP formulation for International T20 India vs Australia (Batting lineup)
To analyze batting and bowling lineups in the cricket world I needed to get the ball-by-ball details
of runs scored by each batsman against each of the bowlers. Fortunately I had already created this with
my R package [yorkr](https://cran.r-project.org/web/packages/yorkr/index.html). yorkr processes yaml data from [Cricsheet](https://cricsheet.org/). So I copied the data of all matches between Australia and India in International T20s. You can download my  processed data for International T20 at [Inswinger](https://github.com/tvganesh/inswinger)



```{r cache=TRUE}
load("Australia-India-allMatches.RData")
dim(matches)
```

The following functions compute the 'Strike Rate' of a batsman as
$$SR=1/overs*\sum RunsScored$$
Also the Economy Rate is computed as
$$ER=1/overs*\sum RunsConceded$$
Incidentally the SR=ER
```{r}
# Compute the Strike Rate of the batsman
computeSR <- function(batsman1,bowler1){
    a <- matches %>% filter(batsman==batsman1 & bowler==bowler1) 
    a1 <- a %>% summarize(totalRuns=sum(runs),count=n()) %>% mutate(SR=(totalRuns/count)*6)
    a1
}

# Compute the Economy Rate of the batsman
computeER <- function(batsman1,bowler1){
    a <- matches %>% filter(batsman==batsman1 & bowler==bowler1) 
    a1 <- a %>% summarize(totalRuns=sum(runs),count=n()) %>% mutate(ER=(totalRuns/count)*6)
    a1
}

```

Here I compute the Strike Rate of Virat Kohli, Yuvraj Singh and MS Dhoni against Shane Watson, Brett Lee
and MA Starc
```{r}
 # Kohli
kohliWatson<- computeSR("V Kohli","SR Watson")
kohliWatson

kohliLee <- computeSR("V Kohli","B Lee")
kohliLee

kohliStarc <- computeSR("V Kohli","MA Starc")
kohliStarc
 

# Yuvraj
yuvrajWatson<- computeSR("Yuvraj Singh","SR Watson")
yuvrajWatson

yuvrajLee <- computeSR("Yuvraj Singh","B Lee")
yuvrajLee

yuvrajStarc <- computeSR("Yuvraj Singh","MA Starc")
yuvrajStarc
 
# MS Dhoni
dhoniWatson<- computeSR("MS Dhoni","SR Watson")
dhoniWatson

dhoniLee <- computeSR("MS Dhoni","B Lee")
dhoniLee

dhoniStarc <- computeSR("MS Dhoni","MA Starc")
dhoniStarc

```

When we consider the batting lineup, the problem is one of maximization.
Formulating and solving 
```{r}
# 3 batsman x 3 bowlers
lprec <- make.lp(0, 9)
# Maximization
a<-lp.control(lprec, sense="max")

# Set the objective function
set.objfn(lprec, c(kohliWatson$SR, kohliLee$SR,kohliStarc$SR,
                   yuvrajWatson$SR,yuvrajLee$SR,yuvrajStarc$SR,
                   dhoniWatson$SR,dhoniLee$SR,dhoniStarc$SR))

#Assume the  bowlers have 3,4,3 overs left respectively
add.constraint(lprec, c(1, 1,1,0,0,0, 0,0,0), "<=",3)
add.constraint(lprec, c(0,0,0,1,1,1,0,0,0), "<=",4)
add.constraint(lprec, c(0,0,0,0,0,0,1,1,1), "<=",3)
#o11+o12+o13+o21+o22+o23+o31+o32+o33=8 (overs remaining)
add.constraint(lprec, c(1,1,1,1,1,1,1,1,1), "=",8) 


add.constraint(lprec, c(1,0,0,0,0,0,0,0,0), ">=",1) #o11 >=1
add.constraint(lprec, c(0,1,0,0,0,0,0,0,0), ">=",0) #o12 >=0
add.constraint(lprec, c(0,0,1,0,0,0,0,0,0), ">=",0) #o13 >=0
add.constraint(lprec, c(0,0,0,1,0,0,0,0,0), ">=",1) #o21 >=1
add.constraint(lprec, c(0,0,0,0,1,0,0,0,0), ">=",1) #o22 >=1
add.constraint(lprec, c(0,0,0,0,0,1,0,0,0), ">=",0) #o23 >=0
add.constraint(lprec, c(0,0,0,0,0,0,1,0,0), ">=",1) #o31 >=1
add.constraint(lprec, c(0,0,0,0,0,0,0,1,0), ">=",0) #o32 >=0
add.constraint(lprec, c(0,0,0,0,0,0,0,0,1), ">=",0) #o33 >=0

lprec
b <-solve(lprec)
get.objective(lprec) #  
get.variables(lprec) # 
```
This shows that the maximum runs that can be scored for the current strike rate is 81.9 runs in 8
overs
The breakup is as follows

Batsman	Watson	B Lee	MA Starc  
	
Kohli	1	1	1  
Yuvraj	0	3	0  
Dhoni	2	0	0  
Overs	3	4	1  
			Total=8  

This is also shown below
```{r}
e <- as.data.frame(rbind(c(1,1,1),c(0,3,0),c(2,0,0),c(3,4,1)))
names(e) <- c("S Watson","B Lee","MA Starc")
rownames(e) <- c("Kohli","Yuvraj","Dhoni","Overs")
e
```

**Note:** This assumes that the batsmen perform at their current Strike Rate. Howvever anything can happen
in a real game, but nevertheless this is a fairly reasonable estimate of the performance


**Note 2:**The numbers in the columns represent the number of overs that need to be bowled by a bowler to the
corresponding batsman.

## 6. LP formulation for International T20 India vs Australia (Bowling lineup)
For this I compute how the bowling should be rotated between R Ashwin, RA Jadeja and JJ Bumrah when
taking into account their performance against batsmen like Shane Watson, AJ Finch and David Warner.
For the bowling performance I take the Economy rate of the bowlers. The data is the same as above

```{r}

computeSR <- function(batsman1,bowler1){
    a <- matches %>% filter(batsman==batsman1 & bowler==bowler1) 
    a1 <- a %>% summarize(totalRuns=sum(runs),count=n()) %>% mutate(SR=(totalRuns/count)*6)
    a1
}
# RA Jadeja
jadejaWatson<- computeER("SR Watson","RA Jadeja")
jadejaWatson

jadejaFinch <- computeER("AJ Finch","RA Jadeja")
jadejaFinch

jadejaWarner <- computeER("DA Warner","RA Jadeja")
jadejaWarner


# Ashwin
ashwinWatson<- computeER("SR Watson","R Ashwin")
ashwinWatson

ashwinFinch <- computeER("AJ Finch","R Ashwin")
ashwinFinch

ashwinWarner <- computeER("DA Warner","R Ashwin")
ashwinWarner

# JJ Bunrah
bumrahWatson<- computeER("SR Watson","JJ Bumrah")
bumrahWatson

bumrahFinch <- computeER("AJ Finch","JJ Bumrah")
bumrahFinch

bumrahWarner <- computeER("DA Warner","JJ Bumrah")
bumrahWarner
```


Formulating solving the bowling lineup is shown below
```{r}
lprec <- make.lp(0, 9)
a <-lp.control(lprec, sense="min")

# Set the objective function
set.objfn(lprec, c(jadejaWatson$ER, jadejaFinch$ER,jadejaWarner$ER,
                   ashwinWatson$ER,ashwinFinch$ER,ashwinWarner$ER,
                   bumrahWatson$ER,bumrahFinch$ER,bumrahWarner$ER))

add.constraint(lprec, c(1, 1,1,0,0,0, 0,0,0), "<=",4) # Jadeja has 4 overs
add.constraint(lprec, c(0,0,0,1,1,1,0,0,0), "<=",3)   # Ashwin has 3 overs left
add.constraint(lprec, c(0,0,0,0,0,0,1,1,1), "<=",4)   # Bumrah has 4 overs left
add.constraint(lprec, c(1,1,1,1,1,1,1,1,1), "=",10) # Total overs = 10
add.constraint(lprec, c(1,0,0,0,0,0,0,0,0), ">=",1)
add.constraint(lprec, c(0,1,0,0,0,0,0,0,0), ">=",0)
add.constraint(lprec, c(0,0,1,0,0,0,0,0,0), ">=",1)
add.constraint(lprec, c(0,0,0,1,0,0,0,0,0), ">=",0)
add.constraint(lprec, c(0,0,0,0,1,0,0,0,0), ">=",1)
add.constraint(lprec, c(0,0,0,0,0,1,0,0,0), ">=",0)
add.constraint(lprec, c(0,0,0,0,0,0,1,0,0), ">=",0)
add.constraint(lprec, c(0,0,0,0,0,0,0,1,0), ">=",1)
add.constraint(lprec, c(0,0,0,0,0,0,0,0,1), ">=",0)

lprec
b <-solve(lprec)
get.objective(lprec) #  
get.variables(lprec) # 
```

The minimum runs that will be conceded by these 3 bowlers in 10 overs is 73.58 assuming the bowling
is rotated as follows


```{r}
e <- as.data.frame(rbind(c(1,0,0),c(2,1,1),c(1,1,3),c(4,2,4)))
names(e) <- c("RA Jadeja","R Ashwin","JJ Bumrah")
rownames(e) <- c("S Watson","AJ Finch","DA Warner","Overs")
e 
#Total overs=10  
```

## 7. LP formulation for IPL (Mumbai Indians - Kolkata Knight Riders - Bowling lineup)
As in the case of International T20s I also have processed IPL data derived from my R package yorkr. [yorkr](https://cran.r-project.org/web/packages/yorkr/index.html). yorkr processes yaml data from [Cricsheet](https://cricsheet.org/). The processed data for all IPL matches can be downloaded from [GooglyPlus](https://github.com/tvganesh/GooglyPlus)

```{r cache=TRUE}
load("Mumbai Indians-Kolkata Knight Riders-allMatches.RData")
dim(matches)
```

```{r}
# Compute the Economy Rate of the batsman

# Gambhir
gambhirMalinga <- computeER("G Gambhir","SL Malinga")
gambhirHarbhajan <- computeER("G Gambhir","Harbhajan Singh")
gambhirPollard <- computeER("G Gambhir","KA Pollard")

#Yusuf Pathan
yusufMalinga <- computeER("YK Pathan","SL Malinga")
yusufHarbhajan <- computeER("YK Pathan","Harbhajan Singh")
yusufPollard <- computeER("YK Pathan","KA Pollard")

#JH Kallis
kallisMalinga <- computeER("JH Kallis","SL Malinga")
kallisHarbhajan <- computeER("JH Kallis","Harbhajan Singh")
kallisPollard <- computeER("JH Kallis","KA Pollard")

#RV Uthappa
uthappaMalinga <- computeER("RV Uthappa","SL Malinga")
uthappaHarbhajan <- computeER("RV Uthappa","Harbhajan Singh")
uthappaPollard <- computeER("RV Uthappa","KA Pollard")


```

Formulating and solving this for the bowling lineup of Mumbai Indians against Kolkata Knight Riders
```{r}
 library("lpSolveAPI")
 lprec <- make.lp(0, 12)
 a=lp.control(lprec, sense="min")
 
 set.objfn(lprec, c(gambhirMalinga$ER, yusufMalinga$ER,kallisMalinga$ER,uthappaMalinga$ER,
                    gambhirHarbhajan$ER,yusufHarbhajan$ER,kallisHarbhajan$ER,uthappaHarbhajan$ER,
                    gambhirPollard$ER,yusufPollard$ER,kallisPollard$ER,uthappaPollard$ER))
 
 add.constraint(lprec, c(1,1,1,1, 0,0,0,0, 0,0,0,0), "<=",4)
 add.constraint(lprec, c(0,0,0,0,1,1,1,1,0,0,0,0), "<=",4)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,1,1,1,1), "<=",4)
 add.constraint(lprec, c(1,1,1,1,1,1,1,1,1,1,1,1), "=",10)
 
 add.constraint(lprec, c(1,0,0,0,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,1,0,0,0,0,0,0,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,1,0,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,1,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,1,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,1,0,0,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,1,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,1,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,1,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,1,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,0,1,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,0,0,1), ">=",0)
 
 lprec
 b=solve(lprec)
 get.objective(lprec) #  
 get.variables(lprec) # 
 
```



```{r}
e <- as.data.frame(rbind(c(3,1,0,0,4),c(0, 1, 0,1,2),c(3, 1, 0,0,4)))
names(e) <- c("Gambhir","Yusuf","Kallis","Uthappa","Overs")
rownames(e) <- c("Malinga","Harbhajan","Pollard") 
e
#Total overs=10  
```


## 8. LP formulation for IPL (Mumbai Indians - Kolkata Knight Riders - Batting lineup)
As I mentioned it is possible to perform a maximation with the same formulation since
computeSR<==>computeER

This just flips the problem around and computes the maximum runs that can be scored for
the batsman's Strike rate (this is same as the bowler's Economy rate)

```{r}
 library("lpSolveAPI")
 lprec <- make.lp(0, 12)
 a=lp.control(lprec, sense="max")
 
 a <-set.objfn(lprec, c(gambhirMalinga$ER, yusufMalinga$ER,kallisMalinga$ER,uthappaMalinga$ER,
                    gambhirHarbhajan$ER,yusufHarbhajan$ER,kallisHarbhajan$ER,uthappaHarbhajan$ER,
                    gambhirPollard$ER,yusufPollard$ER,kallisPollard$ER,uthappaPollard$ER))
 
 
 add.constraint(lprec, c(1,1,1,1, 0,0,0,0, 0,0,0,0), "<=",4)
 add.constraint(lprec, c(0,0,0,0,1,1,1,1,0,0,0,0), "<=",4)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,1,1,1,1), "<=",4)
 add.constraint(lprec, c(1,1,1,1,1,1,1,1,1,1,1,1), "=",11)
 
 add.constraint(lprec, c(1,0,0,0,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,1,0,0,0,0,0,0,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,1,0,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,1,0,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,1,0,0,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,1,0,0,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,1,0,0,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,1,0,0,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,1,0,0,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,1,0,0), ">=",1)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,0,1,0), ">=",0)
 add.constraint(lprec, c(0,0,0,0,0,0,0,0,0,0,0,1), ">=",0)
 lprec
 b=solve(lprec)
 get.objective(lprec) #  
 get.variables(lprec) # 
```


```{r}
e <- as.data.frame(rbind(c(0,3,0,0,3),c(0, 1, 0,3,4),c(0, 1, 3,0,4)))
names(e) <- c("Gambhir","Yusuf","Kallis","Uthappa","Overs")
rownames(e) <- c("Malinga","Harbhajan","Pollard") 
e
#Total overs=11  
```

**Conclusion:** It is possible to thus determine the optimum no of overs to give to a specific bowler
based on his/her Economy Rate with a particular batsman. Similarly one can determine the
maximim runs that can be scored by a batsmen based on their strike rate with bowlers. However,
while this may provide some indication a cricket like any other game depends on a fair
amount of chance.